Asteroseismology {\it sans Fourier}: soon to be crushed by DFM

### 2017-07-05 paper/paper.tex
Tweak parameter fit and add discussion of likelihood

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\tess}{\project{\acronym{TESS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\lsst}{\project{\acronym{LSST}}}

@@

\newcommand{\nukl}{\ensuremath{\nu_{nk\ell}}}
\newcommand{\nukzero}{\ensuremath{\nu_{nk0}}}
\newcommand{\nukone}{\ensuremath{\nu_{nk1}}}
\newcommand{\nuzero}{\ensuremath{\nu_{n00}}}
\newcommand{\dnu}{\ensuremath{\Delta_\nu}}
\newcommand{\numax}{\ensuremath{\nu_{\rm max}}}
\newcommand{\numod}{\ensuremath{{\rm bell}}}
\newcommand{\amps}{\ensuremath{A_{nk\ell}}}
\newcommand{\ampc}{\ensuremath{B_{nk\ell}}}
\newcommand{\rl}{\ensuremath{r_{01}}}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
\received{TBD}
\revised{NA}
\accepted{NA}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the arguement.
\submitjournal{ApJ}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
Stellar oscillations are stochastically driven high-*Q* modes.
If a star is observed through an observing campaign that is shorter
than the coherence time of a mode, that mode will effectively appear
coherent within that time interval.
Here we exploit this to create an inexpensive probabilistic model for
asteroseismology data.
The huge advantage of this approach to traditional asteroseismology
approaches is that it does not require the performance of a fourier
transform or anything like it; this permits the measurement of
asteroseismic parameters in data sets that are irregularly or sparsely
sampled, or in data sets where the critical modes are near the edges
of the available frequency space (near the inverse exposure time at
the high-frequency end or near the inverse of the duration of the
campaign at the low-freqency end).
It also permits the construction of a tractable and justifiable
likelihood function, even when the noise processes are non-trivial.
We produce a parameterized description of an asteroseismic frequency
spectrum and use it along with this likelihood function to produce
posterior estimates of asteroseismic parameters for stars in the
\kepler\ data.
We sub-sample the \kepler\ light curves to (short duration)
\tess-like and (sparsely sampled) \gaia-like data sets and show that
even in these much smaller data sets, we ought to be able to estimate
asteroseismic parameters for many stars, and (because we are just
turning the Bayesian crank) without human intervention.
\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{editorials, notices --- miscellaneous --- catalogs ---
  surveys --- methods of limited utility}

@@

For another example, there is no traditional way to do
asteroseismology with sparse-in-time data, like that which will be
produced by \lsst\ and \gaia.

@@

Finally: Stars are damned coherent. Red giants, in particular, show little
or no sign of decoherence over the entire duration of the \kepler\ Mission.
That should make life better; how?

@@

Consider a set of $2K+1$ modes, indexed by $k \in \mathbb{Z}: -K \le k \le K$. Assuming that the $\ell = 0, 1$ modes are present, we expect a comb of $4K+2$ peaks in the power spectrum. The $\ell = 0$ modes are located at
\begin{equation}
\nukzero = \nuzero (1 + k \dnu), 
\end{equation}
with the $\ell = 1$ modes shifted to
\begin{equation}
\nukone = \nuzero (1 + (k + 1/2) \dnu), 
\end{equation}
Each of these modes corresponds to two sinusoids
\begin{equation}
m(t) = \amps \sin(2\pi\nukl t) + \ampc \cos(2\pi\nukl t)
\end{equation}
into the stellar lightcurve. The amplitudes, $\{\amps, \ampc\}$, are normally distributed with a frequency- and $\ell$-dependent variance:
\begin{equation}
\amps, \ampc \sim \normal \left(\amps, \ampc; 0, \rl^\ell \sigma^2(\nukl) \right),
\end{equation}
where $\rl$, the ratio of the variances of the $\ell = 0$ and $\ell = 1$ modes with equal $k$, is approximately 1/2. The variance, $\sigma^2(\nukl)$, of the amplitudes is modulated by a bell-shaped envelope, yielding
\begin{equation}
\sigma^2(\nu) = \numod(\nu; \numax, W, H),
\end{equation}
where
\begin{equation}
\numod(\nu; \numax, W, H) = H \exp \left( -\frac{(\nu - \numax)^2}{2 W^2} \right).
\end{equation}

This model is completely described by the parameters $\{ \amps, \ampc, \nuzero, \dnu, \rl, \numax, W, H \}$, depending linearly on the amplitudes, $\{\amps, \ampc\}$, and non-linearly on the remainder. In the following, we assume $K=16$, yielding a total of 132 linear parameters and six non-linear parameters.

>>>>

@@

\newcommand{\prob}{\ensuremath{\rm Pr}}
\newcommand{\pars}{\ensuremath{\theta}}
\newcommand{\parsnl}{\ensuremath{\theta_{\rm nl}}}
\newcommand{\parsl}{\ensuremath{\theta_{\rm lin}}}

\newcommand{\nt}{\ensuremath{n_{t}}}
\newcommand{\nlin}{\ensuremath{n_{\rm lin}}}

\newcommand{\nukl}{\ensuremath{\nu_{nk\ell}}}
\newcommand{\nukzero}{\ensuremath{\nu_{nk0}}}
\newcommand{\nukone}{\ensuremath{\nu_{nk1}}}
\newcommand{\nuzero}{\ensuremath{\nu_{n00}}}
\newcommand{\dnu}{\ensuremath{\Delta\nu}}
\newcommand{\numax}{\ensuremath{\nu_{\rm max}}}
\newcommand{\numod}{\ensuremath{{\rm bell}}}
\newcommand{\amps}{\ensuremath{A_{nk\ell}}}
\newcommand{\ampc}{\ensuremath{B_{nk\ell}}}
\newcommand{\eps}{\ensuremath{\kappa_{01}}}
\newcommand{\rl}{\ensuremath{r_{01}}}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
\received{July 1, 2016}
\revised{September 27, 2016}
\accepted{\today}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the arguement.
\submitjournal{ApJ}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

Things, stuff.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{editorials, notices --- 
miscellaneous --- catalogs --- surveys}

@@

For another example, there is no traditional way to do
asteroseismology with sparse-in-time data, like that which will be
produced by LSST and Gaia.

@@

Finally: Stars are damned coherent. Red giants, in particular, show little
or no sign of decoherence over the entire duration of the Kepler mission.
That should make life better; how?

@@

Consider a set of $2K+1$ modes, indexed by $k \in \mathbb{Z}: -K \le k \le K$. Assuming that the $\ell = 0, 1$ modes are present, we expect a comb of $4K+2$ peaks in the power spectrum. The $\ell = 0$ modes are located at
\begin{equation}
\nukzero = \nuzero \, (1 + k \, \Delta), 
\end{equation}
with the $\ell = 1$ modes shifted to
\begin{equation}
\nukone = \nuzero \, (1 + [k + \eps] \, \Delta), 
\end{equation}
with $\eps \simeq 1/2$. Each of these modes corresponds to two sinusoids
\begin{equation}
m_{k\ell}(t) = \amps \, \sin(2\pi \, \nukl \, t) + \ampc \, \cos(2\pi \, \nukl \, t)
\end{equation}
into the stellar lightcurve, yielding $\nlin = 8K+4$ unknown amplitude parameters. The total model lightcurve (i.e., summed over $k$ and $\ell$) can be represented by
\begin{equation}
m(t) = D(t, \nukl) \, \parsl
\end{equation}
where $D$ is an $\nt \, \times \, \nlin$ design matrix and
\begin{equation}
\parsl = \{ A_{n,-K,0}, B_{n,-K,0}, A_{n,-K,1}, B_{n,-K,1}, A_{n,-K+1,0}, B_{n,-K+1,0}, \ldots, A_{n,K,1}, B_{n,K,1} \}
\end{equation}
is a vector of amplitudes. The amplitudes are Gaussian-distributed with a frequency- and $\ell$-dependent variance; for example, 
\begin{equation}
\amps \sim \normal \left(\amps \, ; \, 0, \rl^\ell \, \sigma^2(\nukl) \right),
\end{equation}
where $\rl$, the ratio of the variances of the $\ell = 0$ and $\ell = 1$ modes with equal $k$, is approximately 1/2. The variance, $\sigma^2(\nukl)$, of the amplitudes is modulated by a bell-shaped envelope, yielding
\begin{equation}
\sigma^2(\nu) = \numod(\nu \, ; \, \numax, W, H),
\end{equation}
where
\begin{equation}
\numod(\nu \, ; \, \numax, W, H) = H \, \exp \left( -\frac{[\nu - \numax]^2}{2 \, W^2} \right).
\end{equation}
This model is therefore completely described by the parameter vector $\pars = \{ \amps, \ampc, \nuzero, \Delta, \eps, \rl, \numax, W, H \}$, depending linearly on the amplitudes, $\parsl = \{\amps, \ampc\}$, and non-linearly on the remainder, $\parsnl$.

In the following, we assume $K=16$, yielding a total of 132 linear parameters and seven non-linear parameters. Our aim is to recover the marginal posterior of $\numax$ and $\dnu = \nuzero \, \Delta$, which can be estimated by sampling from the full joint posterior distribution
\begin{equation}
\prob(\pars \, | \, d(t)) \propto \prob(\pars) \, \prob(d(t) \, | \, \pars).
\end{equation}
Na\"ively this involves the daunting task of sampling from a highly multimodal distribution in 140 dimensions. We can, however, greatly reduce the dimensionality by exploiting the linear dependence of the model on the amplitudes. Assuming Gaussian noise with time-domain covariance matrix $C$ yields a Gaussian likelihood function
\begin{equation}
\prob(d(t) \, | \, \pars) = \normal \left(d(t) \, ; \, m(t, \pars), C \right),
\end{equation}
which, coupled with the assumption of Gaussian-distributed amplitudes means we can pre-marginalize over them analytically, yielding
\begin{eqnarray}
\prob(d(t) \, | \, \parsnl) & = & \int d\parsl \, \prob(\parsl) \, \prob(d(t) \, | \, \pars) \nonumber \\
& = & \normal \left(d(t) \, ; \, 0, V(\parsnl) \right).
\end{eqnarray}
Here, the modified covariance matrix $V = C + D \, \Lambda \, D^T$, where $\Lambda$ is a diagonal $\nlin^2$ matrix with non-zero entries 
\begin{equation}
\parsl = \{ \sigma^2(\nu_{0,-K,0}), \sigma^2(\nu_{0,-K,0}), \sigma^2(\nu_{0,-K,1}), \sigma^2(\nu_{0,-K,1}), \ldots, \sigma^2(\nu_{0,K,1}), \sigma^2(\nu_{0,K,1}) \}.
\end{equation}

The slowest steps in evaluating this likelihood are inverting and finding the determinant of the $\nt^2$ covariance matrix $V$; however, provided $\nlin < \nt$ we can increase the speed of this step (potentially significantly: by a factor of $(\nt/\nlin)^3$) by performing low-rank updates of the data covariance: 
\begin{eqnarray}
V^{-1} & = & C^{-1} - C^{-1} \, D \, \left(I + \Lambda \, D^T \, C \, D \right)^{-1} \Lambda \, D^T \, C^{-1} \\
||V|| & = & ||I + \Lambda \, D^T \, C \, D|| \, ||C||.
\end{eqnarray}
Using these forms of the two matrix lemmas is important, as the bell function modulating the variances of the linear amplitudes allows for singular $\Lambda$ matrices, particularly when $W$ is small.
